Type-Safe Modular Parsing

### 2017-09-08 SLE_17/sections/Introduction.tex
Revised

So far research on semantic modularization techniques has focused on
operations that \emph{traverse} or \emph{process} extensible
data structures, such as ASTs. Indeed many documented applications of
semantic modularization techniques focus on modularizing various
aspects of PL implementations.  However, as far as we know, there is
little work on operations that build/produce ASTs.  In particular the
problem of how to modularize parsing has not been studied in semantic
modularization approaches. Techniques such as NOA~\cite{Gouseti2014}
employ a syntactic modularity approach for parsing in combination with
a semantic modularity approach for defining operations
that traverse or process ASTs. This is a shame because parsing is a
fundamental part of PL implementations, and it ought to be made
semantically modular as well, so that the full benefits of semantic
modularity apply.

>>>>

So far research on semantic modularization techniques has focused on
operations that \emph{traverse} or \emph{process} extensible
data structures, such as ASTs. Indeed many documented applications of
semantic modularization techniques focus on modularizing various
aspects of PL implementations.  However, as far as we know, there is
little work on operations that build/produce extensible ASTs, and
guarantee type-safety. 
Sloane and Roberts~\cite{SLOANE201520}
briefly mention a near solution to modular parsing by using case classes. However
the solution could be potentially unsafe since exhaustiveness of
pattern matching for extensible case classes
is not guaranteed. The
problem of how to modularize parsing, including its algorithmic challenges,
has not been studied well-studied yet. Other techniques, such as NOA~\cite{Gouseti2014},
employ a syntactic modularity approach for parsing in combination with
a semantic modularity approach for defining operations
that traverse or process ASTs. Because parsing is a
fundamental part of PL implementations, it ought to be made
semantically modular as well, so that the full benefits of semantic
modularity apply.

### 2017-09-08 SLE_17/sections/RelatedWork.tex
Revised

\paragraph{Modular Parsing with Case Classes}

Our work is close to~\cite{SLOANE201520}, in the sense that they also make use of Packrat parsers
and standard OO inheritance achieving modular parsing. The difference is that case
classes are used in their work for extensibility. Case classes~\cite{emir2007matching} provide a near solution
to the Expression Problem since they can be modular added, and hence are widely used in
practice, but they do not enforce exhaustiveness
of pattern matching for extensible operations. In other words run-time pattern matching errors
can happen when writing extensible code with case classes, so that full static type-safety is not
ensured. Our work uses Object Algebras for full extensibility and type safety, and we have well
studied the algorithmic challenges of parsing in a modular setting.

@@

There are also many design patterns in OO languages that achieve
type-safe extensibility~\cite{torgersen2004expression,odersky2005independently,oliveira2009modular,Oliveira:2012,wang2016expression}. We chose Object Algebras~\cite{Oliveira:2012} because the pattern is
relatively lightweight and makes good use of existing OO features,
such as inheritance, generics and subtyping. As seen throughout the paper,
the parsing code is concise and expressive using Object Algebras.
\vspace{-7pt}

>>>>

@@

There are also many design patterns in OO languages that achieve
type-safe extensibility~\cite{torgersen2004expression,odersky2005independently,oliveira2009modular,Oliveira:2012,wang2016expression}. We chose Object Algebras~\cite{Oliveira:2012} because the pattern is
relatively lightweight and makes good use of existing OO features,
such as inheritance, generics and subtyping. As seen throughout the paper,
the parsing code is concise and expressive using Object Algebras.

Case classes in Scala can encode algebraic datatypes that allow the
addition of new constructors. However such ``open'' case classes 
do not enforce  exhaustiveness
of pattern matching for extensible operations, and thus do not provide 
a full solution to the Expression Problem. Nevertheless case classes 
are widely used in practice, and a solution for parsing open case
classes (and composing such parsers) is quite relevant in practice. 
The techniques in Section~\ref{sec:inheritance} can be readily adapted to work with case classes. 
The work by Sloane and Roberts~\cite{SLOANE201520} on a modular Oberon 
compiler applied similar techniques with packrat parsers and case classes.
In our work we use Object Algebras for full extensibility and type safety, and we have well
studied the algorithmic challenges of parsing in a modular setting.
\vspace{-7pt}
